---
title: "EDUC 152. Intro to quantitative research in education: Regression analysis"
subtitle: "Statistical inference"
author: 
date: 
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true # toc_float option to float the table of contents to the left of the main document content. floating table of contents will always be visible even when the document is scrolled
      #collapsed: false # collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level (e.g., H2) headers. If collapsed initially, the TOC is automatically expanded inline when necessary
      #smooth_scroll: true # smooth_scroll (defaults to TRUE) controls whether page scrolls are animated when TOC items are navigated to via mouse clicks
    number_sections: true
    fig_caption: true # ? this option doesn't seem to be working for figure inserted below outside of r code chunk    
    highlight: tango # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax    
    theme: default # theme specifies the Bootstrap theme to use for the page. Valid themes include default, cerulean, journal, flatly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, and yeti.
    df_print: tibble #options: default, tibble, paged

---


```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

LINK TO GOOGLE DOC WHERE I HAVE WRITTEN OUT TOPICS IN DETAIL

- https://docs.google.com/document/d/1u1RhtIQyohjd87ZVDAm6wH9Pl0EPuYs0zY2sq_3iAaQ/edit 

# Introduction


## Lecture overview

Statistical inference

> The theory, methods, and practice of forming judgments about the parameters of a population and the reliability of statistical relationships, typically on the basis of random sampling 

*credit: [Google](https://www.google.com/search?q=define+%22statistical+inference%22&sxsrf=ALeKk00lHj1hQ5jc8o_Xjsc7vOF6XFMPUg%3A1617292004292&ei=5OplYNCzEdOU-gTl2p_ACQ&oq=define+%22statistical+inference%22&gs_lcp=Cgdnd3Mtd2l6EAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsAMyBwgAEEcQsANQAFgAYPbvD2gBcAJ4AIABvgGIAb4BkgEDMC4xmAEAqgEHZ3dzLXdpesgBCMABAQ&sclient=gws-wiz&ved=0ahUKEwjQtaebst3vAhVTip4KHWXtB5gQ4dUDCA0&uact=5)*

**Lecture overview**

- la la


## Libraries we will use

```{r}
#install.packages('tidyverse') # if you haven't installed already
#install.packages('labelled') # if you haven't installed already

library(tidyverse) # load tidyverse package
library(labelled) # load labelled package package
```
# Introduce IPEDS data

## What/why IPEDS

**What is IPEDS?**

- ADD TEXT

**Which IPEDS variables will we be using to teach statistical inference**

- Variables about annual cost of attendance for full-time graduate programs
  - average tuition (in-state vs. out-of-state)
  - required fees  (in-state vs. out-of-state)
  - books and supplies; off-campus room and board; off-campus "other expenses"
    - these are calculated for undergraduate students, but should be highly correlated with graduate students
  - **Note**: none of these variables include the cost of healthcare
    - Therefore, assuming you must pay for the cost of healthcare, the measures of cost of attendance (COA) we create will understate the total cost of graduate school + life


**Why use IPEDS data rather than *College Scorecard* data? and why these variables?**

- *College Scorecard* brings together variables from many data sources, including IPEDS
  - The most interesting variables in *College Scorecard* are debt and earnings from degree programs, which come from Office of Federal Student Aid and IRS (I think)
- These university-level debt and earnings variables are measured as the mean (or median) of all students who graduated from the university
  - so the underlying data (that we don't see) are student-level; these student-level data are aggregated to the university-level to create measures mean (or median) debt/earnings at that university
  - Using these measures would make teaching statistical inference more confusing (e.g., trying to test hypotheses about the population mean of mean earnings)
- Why teach inferential statistics using IPEDS measures about graduate tuition (and cost of attendance)?
  - These measures are truly measured at the university-level
  - Tuition and cost-of-attendance are the big drivers of student debt, so learning more about how tuition/cost-of-attendance varies across universities will help you make more informed decisions about pursuing graduate education in the future

## Some definitions

Some definitions related to tuition, fees, expenses, etc; from the IPEDS "Student Charges for Full Academic Year" 2019-20 academic year data dictionary [[LINK]](https://nces.ed.gov/ipeds/datacenter/data/IC2019_AY_Dict.zip):

- "Full-time student" (graduate)
  - Graduate - A student enrolled for 9 or more semester credits, or 9 or more quarter credits, or students involved in thesis or dissertation preparation that are considered full time by the institution
- "Academic year"
  - The period of time generally extending from September to June; usually equated to 2 semesters or trimesters, 3 quarters, or the period covered by a 4-1-4 plan
- "Tuition"
  - Amount of money charged to students for instructional services. Tuition may be charged per term, per course, or per credit
- "In-state tuition"
  - The tuition charged by institutions to those students who meet the state's or institution's residency requirements
- "Out-of-state tuition"
  - The tuition charged by institutions to those students who do not meet the state's or institution's residency requirements
- "Required fees"
  - Fixed sum charged to students for items not covered by tuition and required of such a large proportion of all students that the student who does NOT pay the charge is an exception
- "In-state fees"
  - The fees charged by institutions to those students who meet the state's or institution's residency requirements.
- "Out-of-state fees"
  - The fees charged by institutions to those students who do not meet the state's or institution's residency requirements
- "Books and supplies" (undergraduate)
  - Do not include unusual costs for special groups of students (e.g., engineering or art majors), unless they constitute the majority of students at your institution
- "Room charges"
  - The charges for an academic year for rooming accommodations for a typical student sharing a room with one other student. 
- "Board charges"
  - The charge for an academic year for meals, for a specified number of meals per week. 
- "Other expenses"
  - The amount of money (estimated by the financial aid office) needed by a student to cover expenses such as laundry, transportation, entertainment, and furnishings. (For the purpose of this survey room and board and tuition and fees are not included.) 

*Note*: the IPEDS measures of full-time graduate tuition (both in-state and out-of-state) are "average" tuition price across different graduate degree programs (excluding "first-professional" degree programs like law and medicine)

  - e.g., full-time annual tuition price for an MBA program is likely higher than that of an MA in education
  - But in the context of using these measures to teach statistical inference, pretend these prices are not "averages" but rather the official tuition price
  - make the same assumption for variables about fees, books/supplies, living expenses, etc.

*Note*: "required fees" do not include the cost of healthcare (I think)

## Create dataset `df_ipeds_pop`

Load IPEDS dataset

- CONTAINS DATA ON THESE SORTS OF SOURCES
```{r}
load(file = url('https://github.com/anyone-can-cook/educ152/raw/main/data/ipeds/output_data/panel_data.RData'))
```

Create data frame for use in teaching statistical inference
```{r}
df_ipeds_pop <- panel_data %>%
  # keep data from fall 2019
  filter(year == 2019) %>%
  # which universities to keep:
    # 2015 carnegie classification: keep research universities (15,16,17) and master's universities with "larger programs" (18)
    # [REMOVE THIS] also keep: UC-San Francisco; and UC-Hastings College of Law
      #filter((c15basic %in% c(15,16,17,18)) | unitid %in% c(110699,110398)) %>%
  filter(c15basic %in% c(15,16,17,18)) %>%
  # keep publics and private non-profits (i.e., remove for-profits)
  filter(control %in% c(1,2)) %>%
  # which variables to keep
  select(instnm,unitid,opeid6,opeid,control,c15basic,stabbr,city,zip,locale,obereg, # basic institutional characteristics
         tuition6,fee6,tuition7,fee7, # avg tuition and fees for full-time grad, in-state and out-of-state
         isprof3,ispfee3,osprof3,ospfee3, # avg tuition and fees for MD, in-state and out-of-state
         isprof9,ispfee9,osprof9,ospfee9, # avg tuition and fees for Law, in-state and out-of-state
         chg4ay3,chg7ay3,chg8ay3) %>% # [undergraduate] books+supplies; off-campus (not with family) room and board; off-campus (not with family) other expenses
  # rename variables; syntax <new_name> = <old_name>
  rename(region = obereg, # revion
         tuit_grad_res = tuition6, fee_grad_res = fee6, tuit_grad_nres = tuition7, fee_grad_nres = fee7, # grad
         tuit_md_res = isprof3, fee_md_res = ispfee3, tuit_md_nres = osprof3, fee_md_nres = ospfee3, # md
         tuit_law_res = isprof9, fee_law_res = ispfee9, tuit_law_nres = osprof9, fee_law_nres = ospfee9, # law
         books_supplies = chg4ay3, roomboard_off = chg7ay3, oth_expense_off = chg8ay3) %>% # [undergraduate] expenses
  # create measures of tuition+fees
  mutate(
    tuitfee_grad_res = tuit_grad_res + fee_grad_res, # graduate, state resident
    tuitfee_grad_nres = tuit_grad_nres + fee_grad_nres, # graduate, non-resident
    tuitfee_md_res = tuit_md_res + fee_md_res, # MD, state resident
    tuitfee_md_nres = tuit_md_nres + fee_md_nres, # MD, non-resident
    tuitfee_law_res = tuit_law_res + fee_law_res, # Law, state resident
    tuitfee_law_nres = tuit_law_nres + fee_law_nres) %>% # Law, non-resident  
  # create measures of cost-of-attendance (COA) as the sum of tuition, fees, book, living expenses
  mutate(
    coa_grad_res = tuit_grad_res + fee_grad_res + books_supplies + roomboard_off + oth_expense_off, # graduate, state resident
    coa_grad_nres = tuit_grad_nres + fee_grad_nres + books_supplies + roomboard_off + oth_expense_off, # graduate, non-resident
    coa_md_res = tuit_md_res + fee_md_res + books_supplies + roomboard_off + oth_expense_off, # MD, state resident
    coa_md_nres = tuit_md_nres + fee_md_nres + books_supplies + roomboard_off + oth_expense_off, # MD, non-resident
    coa_law_res = tuit_law_res + fee_law_res + books_supplies + roomboard_off + oth_expense_off, # Law, state resident
    coa_law_nres = tuit_law_nres + fee_law_nres + books_supplies + roomboard_off + oth_expense_off) # Law, non-resident    

# Add variable labels to the tuit+fees variables and coa variables
  # tuition + fees variables
    var_label(df_ipeds_pop[['tuitfee_grad_res']]) <- 'graduate, full-time, resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_grad_nres']]) <- 'graduate, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_res']]) <- 'MD, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_nres']]) <- 'MD, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_res']]) <- 'Law, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_nres']]) <- 'Law, full-time, non-resident; avg tuition + required fees'
    
  # COA variables
    var_label(df_ipeds_pop[['coa_grad_res']]) <- 'graduate, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_grad_nres']]) <- 'graduate, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_res']]) <- 'MD, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_nres']]) <- 'MD, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_res']]) <- 'Law, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_nres']]) <- 'Law, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'

df_ipeds_pop %>% glimpse()
```

Show variable labels
```{r}
df_ipeds_pop %>% var_label()
```
Show value labels for variables that are `labelled` class (code note run)
```{r, eval = FALSE}
df_ipeds_pop %>% select(control,locale,region,c15basic) %>% val_labels()
```

## Investigate `df_ipeds_pop`

Investigate data structure

- confirm one observation per unitid
```{r}

df_ipeds_pop %>% group_by(unitid) %>% summarise(n_per_key=n()) %>% ungroup() %>% count(n_per_key)
```

### Tuition/COA at UC campuses

Graduate, state residents
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_grad_res,fee_grad_res,tuitfee_grad_res,books_supplies,roomboard_off,oth_expense_off,coa_grad_res)
```
Graduate, non-resident
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_grad_nres,fee_grad_nres,tuitfee_grad_nres,books_supplies,roomboard_off,oth_expense_off,coa_grad_nres)
```

MD, state resident
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_md_res,fee_md_res,tuitfee_md_res,books_supplies,roomboard_off,oth_expense_off,coa_md_res)
```

MD, non-resident
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_md_nres,fee_md_nres,tuitfee_md_nres,books_supplies,roomboard_off,oth_expense_off,coa_md_nres)
```


Law, state resident
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_law_res,fee_law_res,tuitfee_law_res,books_supplies,roomboard_off,oth_expense_off,coa_law_res)
```

Law, non-resident
```{r}
df_ipeds_pop %>%
  # keep UC campuses
  filter(unitid %in% c(110398,110635,110644,110653,110662,110671,110680,110699,110705,110714,445188,110699,110398)) %>%
  select(instnm,unitid,tuit_law_nres,fee_law_nres,tuitfee_law_nres,books_supplies,roomboard_off,oth_expense_off,coa_law_nres)
```

### Tuition/COA at fancy privates

Some fancy-pants private universities

- USC
  - `unitid == 123961`
- Stanford
  - `unitid == 243744`
- Columbia
  - `unitid == 190150`
- Columbia, Teacher's College
  - `unitid == 196468`
- NYU
  - `unitid == 193900`
- Harvard
  - `unitid == 166027`
- Vanderbilt
  - `unitid == 221999`
- University of Pennsylvania
  - `unitid == 215062`
- Northwestern University
  - `unitid == 147767`
- Johns Hopkins University
  - `unitid == 162928`


Graduate students

- Note: no difference in prices between resident and non-resident
- Note: Some universities did not provide `books_supplies`, `roomboard_off`, `oth_expense_off`
- Some ludicrously unrealistic values for `roomboard_off` and `oth_expense_off`
```{r}
# In-state
df_ipeds_pop %>%
  # keep private fancy pants
  filter(unitid %in% c(123961,243744,190150,196468,193900,166027,221999,215062,147767,162928)) %>%
  select(instnm,unitid,tuit_grad_res,fee_grad_res,tuitfee_grad_res,books_supplies,roomboard_off,oth_expense_off,coa_grad_res)

# Out-of-state
df_ipeds_pop %>%
  # keep private fancy pants
  filter(unitid %in% c(123961,243744,190150,196468,193900,166027,221999,215062,147767,162928)) %>%
  select(instnm,unitid,tuit_grad_nres,fee_grad_nres,tuitfee_grad_nres,books_supplies,roomboard_off,oth_expense_off,coa_grad_nres)
```

MD students
```{r}
df_ipeds_pop %>%
  # keep private fancy pants
  filter(unitid %in% c(123961,243744,190150,196468,193900,166027,221999,215062,147767,162928)) %>%
  select(instnm,unitid,tuit_md_res,fee_md_res,tuitfee_md_res,books_supplies,roomboard_off,oth_expense_off,coa_md_res)
```

Law students
```{r}
df_ipeds_pop %>%
  # keep private fancy pants
  filter(unitid %in% c(123961,243744,190150,196468,193900,166027,221999,215062,147767,162928)) %>%
  select(instnm,unitid,tuit_law_res,fee_law_res,tuitfee_law_res,books_supplies,roomboard_off,oth_expense_off,coa_law_res)
```


### Some basic descriptive statistics

Number of institutions by type
```{r}
# Number of institutions
df_ipeds_pop %>% count()

# Number of institutions by Carnegie type
df_ipeds_pop %>% count(c15basic) %>% as_factor()

# Number of institutions by public/private
df_ipeds_pop %>% count(control) %>% as_factor()

# number of institutions by public/private and carnegie type
df_ipeds_pop %>% count(control,c15basic) %>% as_factor()

# number of institutions by level of urbanization
df_ipeds_pop %>% count(locale) %>% as_factor()

# number of institutions by public/private and level of urbanization
df_ipeds_pop %>% count(control,locale) %>% as_factor()
```

Tuition+fees by public/private and Carnegie type
```{r}
df_ipeds_pop %>% group_by(control,c15basic) %>% 
  summarize(
    sample_size = n(),
    n_nonmiss_tuitfee_res = sum(!is.na(tuitfee_grad_res)),
    mean_tuitfee_res = mean(tuitfee_grad_res, na.rm = TRUE),
    n_nonmiss_tuitfee_nres = sum(!is.na(tuitfee_grad_nres)),
    mean_tuitfee_nres = mean(tuitfee_grad_nres, na.rm = TRUE),    
  ) %>% as_factor()
```

Cost of attendance by public/private and Carnegie type
```{r}
df_ipeds_pop %>% group_by(control,c15basic) %>% 
  summarize(
    sample_size = n(),
    n_nonmiss_coa_res = sum(!is.na(coa_grad_res)),
    mean_coa_res = mean(coa_grad_res, na.rm = TRUE),
    n_nonmiss_coa_nres = sum(!is.na(coa_grad_nres)),
    mean_coa_nres = mean(coa_grad_nres, na.rm = TRUE),    
  ) %>% as_factor()
```

# Distributions

**Definitions**

- Probability distribution

> In probability theory and statistics, a probability distribution is the mathematical function that gives the probabilities of occurrence of different possible outcomes

A little easier to get your head around "frequency distribution" and "relative frequency distribution"

- Frequency distribution
  - for each value of a variable, the number of observations that have that value
- Relative frequency distribution
  - for each value of a variable, the percent of observations that have that value (the number of observations with that value divided by the total number of observations)

We can show frequency distributions (or relative frequency distributions) as a table or as a graph

- Frequency distribution of variable `tuit_grad_res` (for first few values)

```{r}
df_ipeds_pop %>% count(tuitfee_grad_res)
```
- Creating a histogram that shows the frequency distribution of `tuitfee_grad_res`

```{r}
hist(df_ipeds_pop$tuitfee_grad_res, main = 'Frequency distribution of tuitfee_grad_res', freq = FALSE)
#hist(df_ipeds_pop$tuit_grad_res, main = 'Frequency distribution of tuit_grad_res', freq = TRUE)
#lines(density(df_ipeds_pop$tuit_grad_res), col='red', lwd=3)
abline(v = c(mean(df_ipeds_pop$tuitfee_grad_res),median(df_ipeds_pop$tuitfee_grad_res)),  col=c("green", "red"), lty=c(2,2), lwd=c(3, 3))

```
## Normal and skewed distributions

Distributions of variables fall into a few general categories

**Normal distribution**

- a normal distribution is "bell shaped"; has symmetric "tails"
- Below, we generate a variable that has a normal distribution and then plot the variable to visualize what a normal distribution looks like
and then we graph this variable

Create variable with normal distribution

  - variable has a mean of `10`
  - variable ahs a standard deviation of `2`
    - on average, one random observation from the variable is `2` away from the mean
```{r}
num_obs <- 10000

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(num_obs, mean = 10, sd = 2)

mean(norm_dist) # calculate mean
sd(norm_dist) # calculate standard deviation
median(norm_dist) # calculate median
```
Graph the normally distributed variable and plot the mean, median, and standard deviation

```{r}
# ADD PLOT HERE  
```

**Skewed distribution**

- for a variable with a "skewed distribution" one "tail" is longer than the other
- "outliers" are often the cause of skewed distributions
  - outliers are observations that are much higher or much lower than most observations
- skewed distributions are either "left skewed" or "right skewed"

Righ skewed distributions

- the right "tail" is longer than the left due to the presence of positive outliers, defined as observations with very high values compared to most observations
- so there are more positive outliers than you would expect in a bell (normal) shaped variable
- these positive outliers increase the value of the mean, such that the value of the mean is higher than the value of the median
  - mean > median
- real-world variables that tend to be right-skewed
  - income; enrollment size, city population

Create and plot right-skewed variable
```{r}
# ADD CODE
```

  
Left skwed distributions

- the left tail is longer than right tail, usually due to the presence of negative outliers, defined as observations with very low values (e.g., extreme negative values) compared to most observations
- more negative outliers than you would expect in bell shaped variable
- these negative outliers decrease the value of the mean, such that the value of the mean is lower than the value of the median
  - mean < median
- real-world variables that tend to be left-skewed


Create and plot left-skewed variable
```{r}
# ADD CODE
```


```{r}
# https://www.datamentor.io/r-programming/examples/random-number/
# https://stackoverflow.com/questions/28099590/create-sample-vector-data-in-r-with-a-skewed-distribution-with-limited-range

num_obs <- 500

# Generate standard normal distribution (default is mean = 0 and sd = 1)
set.seed(124)
stdnorm_dist <- rnorm(num_obs, mean = 0, sd = 1)  # equivalent to rnorm(10)

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(num_obs, mean = 10, sd = 2)

# Generate left-skewed distribution
set.seed(124)
lskew_dist <- rbeta(num_obs, 5, 2)

# Generate right-skewed distribution
set.seed(124)
rskew_dist <- rbeta(num_obs, 2, 5)

# Create dataframe
df_generated_pop <- data.frame(stdnorm_dist, norm_dist, lskew_dist, rskew_dist)
```

```{r}
# Plot histograms
hist(df_generated_pop$stdnorm_dist, main = 'Standard normal distribution', freq = F)
lines(density(df_generated_pop$stdnorm_dist), col='red', lwd=3)
abline(v = c(mean(df_generated_pop$stdnorm_dist),median(df_generated_pop$stdnorm_dist)),  col=c("green", "red"), lty=c(2,2), lwd=c(3, 3))

hist(df_generated_pop$norm_dist, main = 'Normal distribution', freq = F)
lines(density(df_generated_pop$norm_dist), col='red', lwd=3)
abline(v = c(mean(df_generated_pop$norm_dist),median(df_generated_pop$norm_dist)),  col=c("green", "red"), lty=c(2,2), lwd=c(3, 3))

hist(df_generated_pop$lskew_dist, main = 'Left-skewed distribution', freq = F)
lines(density(df_generated_pop$lskew_dist), col='red', lwd=3)
abline(v = c(mean(df_generated_pop$lskew_dist),median(df_generated_pop$lskew_dist)),  col=c("green", "red"), lty=c(2,2), lwd=c(3, 3))

hist(df_generated_pop$rskew_dist, main = 'Right-skewed distribution', freq = F)
lines(density(df_generated_pop$rskew_dist), col='red', lwd=3)
abline(v = c(mean(df_generated_pop$rskew_dist),median(df_generated_pop$rskew_dist)),  col=c("green", "red"), lty=c(2,2), lwd=c(3, 3))
```

# Sampling distribution

# Hypothesis testing about a population mean

# Fundamentals of causal inference

# Hypothesis testing about two populations
